---
title: "xflex4cast Package: Simulation and Model Fitting Examples"
author: "Mateus Maia et al."
date: "`r Sys.Date()`"
output: github_document 
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,fig.align = 'center')
library(data.table)
library(ggplot2)
library(xflex4cast)  
set.seed(43)
```


# Install and load `xflex4cast`:

```{r, eval=FALSE}
# Install from source
devtools::install("path/to/xflex4cast")

# Load
library(xflex4cast)
```

# Overview

This document demonstrates core simulation experiments and model fitting procedures for the `xflex4cast` package. We present a novel probabilistic framework for forecasting fault counts that captures both typical and extreme events. Non-extreme faults are modeled using a smooth interpolation of multiple additive quantile regressions, while extreme events are described through a discrete generalized Pareto (DGP) distribution. The examples illustrated here correspond to the simulation scenarios of the main manuscript.

## Scenario One: Constant Tail, Covariate in Bulk

### Simulate data

```{r sim_one, fig.align='center'}
sim1 <- simulation_scenario_one(
  n = 5000,
  phi = 0.1,
  xi = 0.3
)

# View summary of simulated data
summary(sim1$data)

# Plot histogram of response variable y
ggplot(sim1$data, aes(x = y)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  theme_minimal() +
  ggtitle("Scenario One: Histogram of Simulated Counts")
```


### Define and Fit the xflex4cast Model
We define the model with formula linking response `y` to covariates `s(x1,k=10)` (true model specification) for the `mqgam` while keeping the scale parameter of the tail model constant. 

```{r fit_first}
mqgam_formula <- y ~ s(x1,k=10)

model1 <- define_xflex4cast(
  mqgam_formula = mqgam_formula,
  mqgam_quantiles = c(0.05, 0.1, 0.25, 0.5, 0.75, 0.8,
                      0.9,0.95,0.99,0.999,0.9999),
  fit_tail = TRUE,
  scale_formula = ~ 1,
  quantile_threshold = 0.9,
  tail_distribution = "DGP"
)

fit1 <- fit(model1, sim1$data)

# Basic summary of tail model 
summary(fit1$tail_mod)

```

### Inspecting Fitted Model Results
```{r, inspecting_model}
# Extract predicted quantiles for original data
predictions1 <- predict(fit1, sim1$data)
print(predictions1[index==1,])

```


## Scenario Two: Tail Scale Varies with Covariates

### Simulate data

```{r sim2, fig.align='center'}

sim2 <- simulation_scenario_two(
  n = 10000,
  phi = 0.1,
  xi = 0.3
)

summary(sim2$data)

ggplot(sim2$data, aes(x = y)) +
  geom_histogram(bins = 30, fill = "tomato", color = "white") +
  theme_minimal() +
  ggtitle("Scenario Two: Histogram of Simulated Counts")

```

### Define and Fit the xflex4cast Model

We define the model with formula linking response `y` to covariates `s(x1,k=10)` for the `mqgam` component and `~x2` to the tail component only. The remaining inspections follow the same structure as the first simulation scenario.

```{r, fit2}
mqgam_formula2 <- y ~ s(x1,k=10) 

model2 <- define_xflex4cast(
  mqgam_formula = mqgam_formula2,
  mqgam_quantiles = c(0.05, 0.1, 0.25, 0.5, 0.75, 0.8 ,0.9,
                      0.95,0.99,0.999,0.9999),
  fit_tail = TRUE,
  scale_formula = ~ x2,
  quantile_threshold = 0.9,
  tail_distribution = "DGP"
)

```

